SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-05 paper.tex
edits

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian, if naively implemented, scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., for direct numerical simulations, evaluation of reaction rates with a 22-species model can consume around half of the total runtime~\cite{Spafford:2010aa}.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most common implicit integration techniques utilized to deal with numerical stiffness---becomes a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, and necessitates use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques exist to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, falling broadly into a few categories; removal of unimportant species and reactions unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce chemical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.
In addition to the previously mentioned cost reduction methods, effort has gone into improving the integration algorithms and codes that evaluate the chemical kinetics~\cite{Gou:2010,SCHWER2002270,Niemeyer:2016aa,GAO2015287}.
In particular, a carefully derived analytical formulation of the Jacobian can greatly increase the sparsity of the matrix~\cite{SCHWER2002270} and also drop the cost of Jacobian evaluation from scaling quadratically with the number of species in the model---for a naive finite-difference implementation---to a linear dependence~\cite{LU2009192}, while Jacobian factorization can be accelerated via sparse-matrix techniques~\cite{superlu99}.
Use of Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processors to accelerate chemical kinetic simulations has also been studied, e.g., as in~\cite{CurtisGPU:2017,Sewerin20151375,Shi:2012aa,Niemeyer:2016aa}.

@@

A typical modern CPU has many compute units (i.e., cores), each with specialized vector processing units capable of running SIMD instructions (\cref{F:SIMD}).
A SIMD instruction utilizes the vector processor to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently; the number of concurrent operations possible is known as the vector-width\footnote{OpenCL allows for use of vector-widths different from the actual hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied in Sec.~\ref{S:results}} and is typically around 2--4 double precision operations. 
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (MIC) have been developed that have tens of cores with very wide vector-widths (e.g. 4--8 double precision operations); these very wide vector-widths are also found on cutting-edge and expected on forthcoming Intel CPUs (the Skylake Xeon and Cannon Lake architectures).

@@

Recognizing the bottleneck caused by evaluation and factorization of the chemical kinetic Jacobian, a number of recent works have developed analytical chemical kinetic Jacobian codes, although as will be discussed at the end of this section this work offers several improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including incompatibilities with modern reaction types (i.e., pressure dependent Arrhenius and Chebyshev reactions), and is not safely run in parallel~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} discussed the accuracy of Jacobian matricies evaluated with finite-differences compared to analytical derivations and their affect on computational efficiency as well as on the accuracy of time-scale analysis techniques such as CSP~\cite{Lam:1994ws}.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; an \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, supporting both the CPU and GPU, finding speedups of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it is not publicly available nor was it validated (outside the context of use with an implicit-integration technique); in addition the Jacobian was based on an over-constrained system, and a differential algebraic solver should be used for strict correctness.

@@

Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050; paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells, an \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.

@@

Niemeyer et al.~\cite{Niemeyer:2016aa} additionally developed and validated an analytical chemical kinetic Jacobian code for the CPU and GPU.
Yonkee and Sutherland~\cite{Yonkee2016} implemented SIMT-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and a Tesla K20 GPU.

@@

Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two-fourth order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to VODE running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two smaller chemical models with an integration time step of \SI{e-6}{$\sec$}.

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock atmospheric chemical kinetics solver on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on the CBE finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (i.e., \texttt{\#pragmas}).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
To our knowledge, this is the first work that provides vectorized source term and Jacobian evaluation for any chemical model on the CPU, GPU and other accelerators.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with K-rows and N-columns and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example the concentration of species $j$ for the $i$-th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$, the number of thermo-chemical states considered, and $1 \le j \le \ns/$, the number of species in the model.
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2} etc. as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates thermo-chemical source terms\slash chemical kinetic Jacobian for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{L, j}$, the concentration of species $j$ for states $1, 2,\ldots L$---the size of the vector lane for SIMD or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration however, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{NVIDIA:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source terms for a single thermo-chemical state, thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + L}$---the species concentrations for state 1, species $j \ldots j + L$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two species concentrations within the same thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but leads to uncoalesced memory-accesses on the GPU.
However, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads, and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g. resulting from different if\slash then branches)

@@

Finally,~\cref{F:vector_data} shows a vectorized data-ordering that improves the caching patterns of a shallow, ``C''-ordered SIMD vectorization on the CPU (\cref{F:row_major_split}) and a deep, ``F''-ordered SIMT vectorization on the GPU (\cref{F:column_major_split}).
This is accomplished by splitting the slower-varying axis of the data array---rows for ``C''-ordering, and columns for ``F''-ordering---into chunks of size $L$ (the SIMD vector-width, or SIMT warp-size) and laying these data out contigously in memory.
For example, using the data-ordering in~\cref{F:row_major_split}, the concentration of species $j$ for states $i\ldots i+L$, $[C]_{i, j} \ldots [C]_{i + L, j}$, are contiguous in memory and are followed by the concentration of species $j + 1$ for the same states, $[C]_{i, j + 1} \ldots [C]_{i + L, j + 1}$.
This pattern, similar to OpenCL's native vector data-types---e.g., \texttt{double\num{8}} which treats \num{8} contiguous double precision floating point numbers as a single vector datum---ensures that any SIMD operation occurs on data that is contiguous in memory, greatly improving caching and SIMD-throughput.
Conversely, the data-ordering in~\cref{F:column_major_split} is designed to enable coalesced memory accesses for ``F''-ordered, deep-SIMT vectorization on the GPU.
The effects of these various data-ordering and vectorization patterns on performance will be studied in~\cref{S:results}.

\subsection{Thermo-chemical Source Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to our complete derivations, attached as supplemental material to this document.

The thermo-chemical state vector derivation consists of the temperature, the non-constant thermodynamic state parameter (volume or pressure) and the number of moles of all species except the last in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns/$ total species).

This state vector---inspired by~\cite{SCHWER2002270}---has a number of features that recommend its use.
First, the state vector results in highly sparse chemical kinetic Jacobians as will be detailed in~\cref{S:sparsity}.
Second, mass is explicitly conserved in this formulation owing to the fact that the moles of the final species is calculated from the ideal gas law, and the rate of change of the final species from conservation of mass (see \cref{deriv-s:state,deriv-s:source} of the supplemental material respectively); i.e., system is not over-constrained.
Finally, the chemical kinetic Jacobian changes relatively little between the \conp/ and \conv/ forms, making maintaining the code-base much simpler.
Although most current combustion codes do not use species moles as a state variable, to\slash from more common mass\slash mole-fractions and moles is straightforward, and once inside integration of an chemical kinetic initial value problem, the choice of variables no longer matter.

@@

For both \conp/ and \conv/, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\loadeq{dmolar}
where $\dot{\omega}_k$ is the $kth$ species' overall production rate:
\begin{equation}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr/} \nu_{k, i} R_{i} c_{i}
\end{equation}
where $\nu_{k, i}$ is the net stoichiometic coefficient of species $k$ in reaction $i$, $R_{i}$ the net rate of progress of reaction $i$ (described in~\cref{deriv-s:rop_net} of the supplemental), and $c_{i}$ the pressure modification term i.e., for third-body (see~\cref{deriv-s:thdbody} of the supplimental), or fall-off\slash chemically-activated reactions (\cref{deriv-s:falloff} of the same).

@@

The molar source terms are more complicated due to the derivatives of the net rate of progress---see~\cref{deriv-s:dropi}---and pressure modification terms---\cref{deriv-s:dci}---for the reactions.
The reader is again encouraged to read the supplimental material covering the exact derivations and forms of the reactions solved by \texttt{pyJac}.
The derivative of the molar source term with respect to temperature is:
\loadeq{dnkdotdT}
while, similarly the derivative with respect to the state parameter is:
\loadeq{dnkdotdE}
and the moles of species $j$:
\loadeq{dnkdotdnj}

>>>>

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total runtime.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most common implicit integration techniques utilized to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, and necessitates use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques exist to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, falling broadly into a few categories; removal of unimportant species and reactions unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce chemical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.
In addition to the previously mentioned cost reduction methods, effort has gone into improving the integration algorithms and codes that evaluate the chemical kinetics~\cite{Gou:2010,SCHWER2002270,Niemeyer:2016aa,GAO2015287}.
In particular, a carefully derived analytical formulation of the Jacobian can greatly increase the sparsity of the matrix~\cite{SCHWER2002270} and drop the cost of Jacobian evaluation from scaling quadratically with the number of species in the model---for a naive finite-difference implementation---to a linear dependence~\cite{LU2009192}, while Jacobian factorization can be accelerated via sparse-matrix techniques~\cite{superlu99}.
Use of Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processors to accelerate chemical kinetic simulations has also been studied, e.g., as in~\cite{CurtisGPU:2017,Sewerin20151375,Shi:2012aa,Niemeyer:2016aa}.

@@

A typical modern CPU has many compute units (i.e., cores), each with specialized vector processing units capable of running SIMD instructions (\cref{F:SIMD}).
A SIMD instruction utilizes the vector processor to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently; the number of concurrent operations possible is known as the vector-width\footnote{OpenCL allows for use of vector-widths different from the actual hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied in Sec.~\ref{S:results}} and is typically around 2--4 double precision operations. 
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (MIC) have been developed that have tens of cores with very wide vector-widths (e.g. 4--8 double precision operations); these very wide vector-widths are also found on cutting-edge and forthcoming Intel CPUs; i.e., the Skylake Xeon and Cannon Lake architectures, respectively.

@@

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types (i.e., pressure dependent Arrhenius and Chebyshev reactions), and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, and a differential algebraic equation (DAE) solver should be used for strict correctness\slash conservation.

@@

Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050; paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells; a \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.

@@

Yonkee and Sutherland~\cite{Yonkee2016} implemented GPU-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and a Tesla K20 GPU.

@@

Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two-fourth order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{cvode:2.8.2} running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two smaller chemical models with an integration time step of \SI{e-6}{$\sec$}.

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock integrator for atmospheric chemical kinetics on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a SIMD-vectorized third-order stiff Rosenbrock integrator for atmospheric chemistry on the CBE finding a speedup of \SI{1.89}{$\times$} (a parallel scaling efficiency of \SI{0.94}{$\percent$}) over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
SIMD-vectorization improved the integrator performance over an OpenMP baseline (which was vectorized by simple compiler hints (i.e., \texttt{\#pragmas}) by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
To our knowledge, this is the first effort that enables vectorized chemical-kinetic source-term and Jacobian evaluation for any chemical model on the CPU, GPU and other accelerators.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with K-rows and N-columns and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example the concentration of species $j$ for the $i$-th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$, the number of thermo-chemical states considered for evaluation, and $1 \le j \le \ns/$, the number of species in the model.
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2} etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates thermo-chemical source terms\slash chemical kinetic Jacobian for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{N_v, j}$---the concentration of species $j$ for states $1, 2,\ldots N_v$, where $N_v$ is the size of the vector lane for SIMD or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{NVIDIA:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + L$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two species concentrations within the same thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches)

@@

Finally,~\cref{F:vector_data} shows a vectorized data-ordering that improves the caching patterns of a shallow, ``C''-ordered SIMD vectorization on the CPU (\cref{F:row_major_split}) and a deep, ``F''-ordered SIMT vectorization on the GPU (\cref{F:column_major_split}).
This is accomplished by splitting the slower-varying axis of the data array---columns for ``C''-ordering, and rows for ``F''-ordering---into chunks of size $N_v$ (the SIMD vector-width, or SIMT warp-size) and laying these data out contiguously in memory.
For example, using the data-ordering in~\cref{F:row_major_split}, the concentration of species $j$ for states $i\ldots i+N_v$, $[C]_{i, j} \ldots [C]_{i + N_v, j}$, are contiguous in memory and are followed by the concentration of species $j + 1$ for the same states, $[C]_{i, j + 1} \ldots [C]_{i + N_v, j + 1}$.
This pattern, similar to OpenCL's native vector data-types---e.g., \texttt{double\num{8}} which treats \num{8} contiguous double precision floating point numbers as a single vector datum---ensures that any SIMD operation occurs on data that is contiguous in memory, greatly improving caching and SIMD-throughput.
Conversely, the data-ordering in~\cref{F:column_major_split} is designed to enable coalesced memory accesses for ``F''-ordered, deep-SIMT vectorization on the GPU.
The effects of these various data-ordering and vectorization patterns on performance will be studied in~\cref{S:results}.

\subsection{Thermo-chemical Source Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to the complete derivations, attached as supplemental material to this document.

The thermo-chemical state vector derivation consists of the temperature, a non-constant thermodynamic state parameter (volume or pressure for \conp/ and \conv/ respectively), and the number of moles of all species except the last species in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns/$ total species).

This state vector---inspired by~\cite{SCHWER2002270}---has a number of features that recommend its use.
First, the state vector results in highly sparse chemical kinetic Jacobians as will be detailed in~\cref{S:sparsity}.
Second, mass is explicitly conserved in this formulation owing to the fact that the moles of the final species is calculated from the ideal gas law, and the rate of change of the final species from conservation of mass (see \cref{deriv-s:state,deriv-s:source} of the supplemental material respectively); additionally, system is not over-constrained and does not require use of a more complicated DAE solver (as compared to an regular ODE solver) for integration\todo{hansen citation here}.
Finally, the chemical kinetic Jacobian changes relatively little between the \conp/ and \conv/ forms, making maintaining the code-base much simpler.
Although most current combustion codes do not use species moles as a state variable, conversion to\slash from more common mass\slash mole-fractions and moles is straightforward, and once inside integration of an chemical kinetic initial value problem (IVP), the choice of variables no longer matter.

@@

For both \conp/ and \conv/, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\loadeq{dmolar}
where $\dot{\omega}_k$ is the $kth$ species' overall production rate:
\begin{equation}
 \dot{\omega}_{k} = \sum_{i=1}^{\nr/} \nu_{k, i} R_{i} c_{i}
\end{equation}
with $\nu_{k, i}$ the net stoichiometic coefficient of species $k$ in reaction $i$, $R_{i}$ the net rate of progress of reaction $i$ (described in~\cref{deriv-s:rop_net} of the supplemental), and $c_{i}$ the pressure modification term i.e., for third-body (see~\cref{deriv-s:thdbody} of the supplemental), or fall-off\slash chemically-activated reactions (\cref{deriv-s:falloff} of the same).

@@

The molar source terms are more complicated due to the derivatives of the net rate of progress---see~\cref{deriv-s:dropi}---and pressure modification terms---\cref{deriv-s:dci}---of the reactions.
The reader is again encouraged to read the supplemental material covering the exact derivations and forms of the reactions solved by \texttt{pyJac}.
The derivative of the molar source term with respect to temperature is:
\loadeq{dnkdotdT}
while, similarly the derivative with respect to the state parameter is:
\loadeq{dnkdotdE}
and the moles of species $j$:
\loadeq{dnkdotdnj}

\subsection{Code Generation and Testing Infrastructure}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL\slash C-code.
This allows for easy changes of program structure, e.g. data ordering, vectorization, threading patterns and switching of target-languages (and simpler extension to additional languages, e.g., CUDA).
Additionally, \texttt{loo.py} can execute developed subroutines from python (natively for C-code, or via \texttt{PyOpenCL}~\cite{kloeckner_pycuda_2012} for OpenCL-code), enabling unit-testing\slash validation for each component of the Jacobian or source terms.
The source-terms and subcomponents thereof (e.g., rates of progress, pressure-modification terms, etc.) are directly compared to Cantera~\cite{Cantera}, while the automatic differentiation code Adept~\cite{adept-v11} is used to obtain reference answers for Jacobian subcomponents.
The Portable OpenCL (POCL) implementation~\cite{poclIJPP} and OpenMP~\cite{dagum1998openmp} are used to run OpenCL and C unit-testing, respectively, on the continous integration framework Travis CI~\cite{travis:2018}.
Validation of the complete (as opposed to the subcomponent testing discussed here) generated source-terms and Jacobian codes will be discussed in detail in~\cref{s:validation}.
